---
title: "DataTrans_part1_2017-05-04.Rmd"
author: "Stacey Harmer"
date: "5/4/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Transformation (Ch 3 in book, Ch 5 online)

```{r}
library(nycflights13)
library(tidyverse)
flights
```

#### 5.2  Filter rows with filter
This allows you to subset based on values.
```{r}
jan1 <- filter(flights, month == 1, day == 1)
near(sqrt(2)^2, 2) # kind of neat
```
###### comparisons
remember 'near'

###### logical operators
remember 'is.na' 
```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)
filter(df, is.na(x) | x > 1)
```

##### 5.2.4  Exercises

##### 5.2.4.1  Find all flights that:
```{r}
# 1.  had arrival delay > 2 hrs

(delay <- filter(flights, arr_delay > 120))
summary(delay)

# 2. flew to IAH or HOU
(hou <- filter(flights, dest == "IAH" | dest == "HOU" ))

# 3.  Operated by United, America, Delta
(carr <- filter(flights, carrier == "UA"| carrier == "AA" | carrier == "DL"))
summary(carr) # note that NAs are still included

summary(carr) # note that NAs are still included
(carr2 <- filter(flights, !is.na(dep_delay), carrier == "UA"| carrier == "AA" | carrier == "DL"))
summary(carr2) #NAs not in this tibble

# 4. departed in summer
(summ <- filter(flights, month == 7 | month == 8 | month == 9))

# 5. left without delay, but arrived more than 2 hrs late
(slow <- filter(flights, dep_delay <= 0 & arr_delay > 120))

# 6. delayed at least 1 hr, but arrived < 30 min late
(fast <- filter(flights, dep_delay >= 60 & arr_delay <= 30))

# 7. departed between midnight and 6 am
(late <- filter(flights, dep_time <= 0600 | dep_time == 2400))
```

##### 5.2.4.2  What does between() do?
```{r}
?between
between(1:10, 4, 6)  # returns list of T/F 

# simplify any of the above code?
# yes, number 4

(summ2 <- flights[between(flights$month, 7, 9) ,])

# or, 
summ3 <- filter(flights, between(month, 7, 9))

library(compare)
compare(summ, summ2)  #TRUE.  so this did work as expected
compare(summ, summ3)  # TRUE also
```

##### 5.2.4.3  How many flights have a missing dep_time?
```{r}
summary(flights)
# 8255 flights don't list a departure time, although they did all have scheduled dep_times
# same number (8255) also lack dep_delay; same flights.  
# arr_time missing for 8713; arr_delay missing for 9430
# I'm not sure why these are different.  

# anyway, presumably some flights never took off (8255) and then additional flights didn't make it to their final destination

```

##### 5.2.4.4  NA
```{r}
NA ^ 0 # 1; because any positive number to 0 = 1 (but what about negative?)
-5 ^ 0
NA | TRUE  #True; that makes sense (| means or, after all)
FALSE & NA # False.  why not NA?  well, can't be TRUE.
# if something is False, it is False no matter other attributes

NA * 0  # NA.  why not 0??  
 -1 * 0 # 0
Inf * 0 # NaN
#  so x * 0 is not always 0
```

#### 5.3  Arrange rows with arrange
Changes order of rows.  usage seems pretty intuitive
```{r}
arrange(flights, year, month, day)
```
Can stick 'desc()' in there to reverse order of sorting

##### 5.3.1  Exercises

###### 5.3.1.1  use arrange() to sort all NAs to the start
```{r}
sort.NA.top <- arrange(flights, desc(is.na(arr_delay)))
head(sort.NA.top) # seemed to work
```

###### 5.3.1.2.  sort flights to find most delayed flights
```{r}
(del.flight <- arrange(flights, desc(dep_delay)))
head(del.flight)
# "find the flights that left the earliest"
# not sure what that means. The least delayed flights?

```

##### 5.3.1.3.  Sort flights to find fastest flights
```{r}
?flights
# this might mean find the shortest flights
short.flights <- arrange(flights, air_time)
head(short.flights)

# or, might mean fastest rate of travel (distance/air_time)
flights.rate <- flights
flights.rate$mpm <- flights$distance / flights$air_time

fast.flights <- arrange(flights.rate, desc(mpm))

```
Fastest flight is 11.7 miles per minute, or 702 miles per hour

###### 5.3.1.4.  longest and shortest flights
```{r}

summary(flights)
# min is 17 ; max is 4982 miles
```


#### 5.4  Select columns with select
Some regular expressions are possible here
```{r}
select(flights, starts_with("c"))
select(flights, ends_with("e"))

```
rename() looks very useful
```{r}
str(iris)
# rename(iris, Species = species) # failed. Error: Unknown variables: species.
rename(iris, species = Species) # that worked! new name first, previous name second
str(flights)
tailnum <- rename(flights, tail_num = tailnum) # that worked
str(tailnum) # now 'tailnum' replaced by 'tail_num'

select(flights, arr_time, dep_time, everything()) # this worked
```

##### 5.4.1  Exercises

###### 5.4.1.1. Think of lots of ways to select dep_time, dep_delay, arr_time, and arr_delay

```{r}
dep_arr <- c("dep_time", "arr_time", "dep_delay", "arr_delay")
select(flights, one_of(dep_arr)) # that worked

select(flights, dep_time, dep_delay, arr_time, arr_delay)
# that worked

select(flights, starts_with("dep"), starts_with("arr"))
# that worked

str(flights) #these are columns 4, 6, 7, 9
select(flights, c(4,6,7,9)) # that worked

select(flights, contains("e")) # too many columns; didn't work!

```

###### 5.4.1.2. include name of avariable multilple times in select()
```{r}
select(flights, dep_delay, dep_delay)
select(flights, dep_delay) # these seem the same; redundant calls ignored
```

###### 5.4.1.3. what does one_of() funciton do?
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

select(flights, one_of(vars)) # I got all 5 columns; returns columns that match one of above
```
Note that this allows us to deal with character vector, not just unquoted variable names.


###### 5.4.1.4.  run code and explain 
```{r}
select(flights, contains("TIME"))
# I am surprised; it ignored case.  

select(flights, contains("TIME", ignore.case = FALSE))
# but now it doesn't
```

#### 5.5  Mutate - add new variables
```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
head(flights_sml)

mutate(flights_sml,
  gain = arr_delay - dep_delay,
  speed = distance / air_time * 60
)
```
interestingly, function transmute() just puts products of new operation into new tibble
```{r}
transmute(flights,
  gain = arr_delay - dep_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

#### 5.5.1  Useful creation functions
Lots of ways to create new columns/variables.
Remember recycling rule for parameter length

modular arithmetic: %% indicates x mod y and %/% indicates integer division,
or "x %% y	means _remainder of x divided by y (x mod y)_" 
and "x %/% y	means _x divided by y but rounded down (integer divide)_"

Offsets - lead() and lag()
```{r}
x <- 1:10
lag(x)
lead(x)  
# note the NA within the output
```
cumulative and rolling aggregates
```{r}
x
cumsum(x)
cummean(x)
```

ranking
```{r}
y <- c(1, 2, 2, NA, 3, 4)
min_rank(y) # is NA ignored, and the 2 '2' values called 2 and 3?

ntile(y, 2) # seems like it returns '1' when number in vector is <= 2, but '2' when > 2
ntile(y, 3)
```
I don't quite understand min_rank?
row_number() doesn't generate ties the way min_rank() does.
dense_rank() allows ties, but those don't count towards final list of numbers
percent_rank() sets range, with min at 0 and max at 1.0
cume_dist() sets range from 0 to 1, and each point as fraction along that range

ntile : Finally, ntile() divides the data up into n evenly sized buckets. It’s a coarse ranking, and it can be used in with mutate() to divide the data into buckets for further summary. 
ntile() is weird.  seems to be whether 

#### 5.5.2  Exercises

##### 5.5.2.1  Convert dep_time and sched_dep_time to # minutes since midnight
```{r}
head(flights)
flights_minutes <- transmute(flights, 
                             dep_time_min = dep_time %/% 100 * 60 + dep_time %% 100,
                            sched_dep_time_min = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100)
head(flights_minutes)
head(flights)
# working out the format
test <- c(1215, 905, 620)
(hour <- test %/% 100)
(min <- test %/% 100 * 60)
(rem <- test %% 100)
```

##### 5.5.2.2  Compare air_time with arr_time - dep_time.
Naively, I'd expect the two to be equal
```{r}
flights_2 <- select(flights, 
   year:day,                 
  contains("time"),
  -time_hour
)
head(flights_2)

flights_2 <- mutate(flights_2, comp_travel_time = arr_time - dep_time)
# rearrange for ease of comparison

flights_2 <- arrange(flights_2, air_time, comp_travel_time)
# seems that the computed travel time is always greater.
# let's graph that to be sure

flights_2 <- mutate(flights_2, wait_time = comp_travel_time - air_time)
summary(flights_2)

ggplot(flights_2, aes(comp_travel_time, air_time)) +
  geom_jitter(aes(alpha = 0.2))
```
Hmm, there are two distinct groups.
Often the differnece between 'air_time' and 'arr_time - dep_time' is small.
In these cases, computed values always positive (computed = longer than air)
This may sometimes be time spent on the tarmac.

Sometimes, computed travel time is much less than air_time.  A fault of the time display?
Try again, with minutes rather than hours for computed times

```{r}

flights_minutes_2 <- transmute(flights, 
                             arr_time_min = arr_time %/% 100 * 60 + arr_time %% 100,
                            dep_time_min = sched_dep_time %/% 100 * 60 + dep_time %% 100, 
                            comp_time_min = arr_time_min - dep_time_min, 
                            air_time = air_time)
head(flights_minutes_2)
 # now look at the difference between computed and reported airtime

flights_minutes_2 <- mutate(flights_minutes_2, wait_time = comp_time_min - air_time)
# and now plot out
ggplot(flights_minutes_2, aes(comp_time_min, air_time)) +
  geom_jitter(aes(alpha = 0.2))
# still some wonky values, with cmoputed air times much less than reported.
# 24 hours = how many minutes?
60*24 # 1440.  Maybe that explains it.  Next day arrival.

ggplot(flights_minutes_2, aes(wait_time, air_time)) +
  geom_jitter(aes(alpha = 0.2))
# and some negative values for wait time.  that is hard to understand.
# Wait, maybe time zone issues? 
```
I would need to know the differences between arrival and departure time zones to fix the data

##### 5.5.2.3  Compare dep_time, sched_dep_time, and  dep_delay.  How do you expect them to be related?

I expect dep_time - sched_dep_time == dep_delay.  let's check
```{r}

flights_3 <- select(flights, 
   year:day,                 
  contains("dep")
)
head(flights_3)

flights_3 <- mutate(flights_3, 
                    dep_time_mins = dep_time %/% 100 * 60 + dep_time %% 100, 
                    sched_dep_time_mins = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100
                    )

flights_3 <- mutate(flights_3, calc_dep_delay = dep_time_mins - sched_dep_time_mins)
summary(flights_3)
# min is -1438; max is 911.  Large negatives meant delay until next day, I would think?

ggplot(flights_3, aes(calc_dep_delay, dep_delay)) +
  geom_point(aes(alpha = 0.2))

# OK, two linear plots!  The one to the left offset by about 60*24 =   1440 minutes. 
# so next day departure
```

##### 5.5.2.4  Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank()

OK, I want to rank by dep_delay.  "most common ranking is with min_rank"
That is, tied values are all given the lowest rank
```{r}

summary(flights) # max dep_delay is  1301.00
flights_rank <- select(flights, year:day, dep_time, sched_dep_time, dep_delay, carrier, flight)
head(flights_rank)

flights_rank <- mutate(flights_rank, delay_rank = min_rank(desc(flights$dep_delay)))
# now order them by that new column

flights_rank <- arrange(flights_rank, delay_rank, carrier, flight)
head(flights_rank, n = 10)
# no ties inthe top 10.    
# HA is Hawaiian Air, MQ is Envoy Air, AA is American, DL is Delta.  
?flights
airlines  #DL has 4 of top 10; MQ has 3.  
```


##### 5.5.2.5  What does 1:3 + 1:10 return? Why?


```{r}
#  1:3 + 1:10

```
Error, because vectors of different lenghts. Try a few others
```{r}
1:5 + 1:10  # worked.  vector of length 10.  cycled through 1:5 twice

1:10 + 1:5 # same values as above

1:5 + 2 # that is more intuitive

```

##### 5.5.2.6  What trig functions does R provide?

```{r}
?Trig
```
These functions give the obvious trigonometric functions. They respectively compute the cosine, sine, tangent, arc-cosine, arc-sine, arc-tangent, and the two-argument arc-tangent.



