---
title: "DataExp_part2_2017-05-28.Rmd"
output: 
  html_document: 
    keep_md: yes
---
chapter 7, part 2 (7.6 - 7.8)
Workflow chapter
Tibbles chapter

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 7.  Data Exploration 

### 7.6 patterns and models

```{r}
library(tidyverse)
library(modelr)
mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid))

ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))

ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))

```

### 7.7 ggplot calls

```{r}
diamonds %>% 
  count(cut, clarity) %>% 
  ggplot(aes(clarity, cut, fill = n)) + 
    geom_tile()  # can pipe the extracted data to gglot without explicitly naming it
```

## 8.  Workflow

keyboard shortcuts that will work together to make sure you’ve captured the important parts of your code in the editor:

Press Cmd/Ctrl + Shift + F10 to restart RStudio.
Press Cmd/Ctrl + Shift + S to rerun the current script.

### 8.3 Paths

Never use absolute paths in scripts - bad for sharing.

```{r}
getwd()

```

## 10.  Tibbles

### 10.2  Creating Tibbles

```{r}
as_tibble(iris)

tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)

ex <- tribble(
  ~x,  ~y,
  "a", 1:3,
  "b", 4:6
)
ex
ex$y  # now I can see the lists

```

Tibbles can handle non-standard column names. Just use the backtick, which is found on my keyboard on same key as ~

```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
```

A Tribble is a transposed tibble

```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)

nycflights13::flights %>% 
  print(n = 10, width = Inf)
```

### 10.3  Subsetting

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
#> [1] 0.434 0.395 0.548 0.762 0.254
df[["x"]]
#> [1] 0.434 0.395 0.548 0.762 0.254

# Extract by position
df[[1]]


```

 NOTE - to use subsetted data in a pipe, you need a placeholder  (a period)
 
 
```{r}
df %>% .$x
#> [1] 0.434 0.395 0.548 0.762 0.254
df %>% .[["x"]]

```

### 10.5  Exercises

#### 10.5.1 How can you tell if an object is a tibble?
```{r}
ex <- tribble(
  ~x,  ~y,
  "a", 1:3,
  "b", 4:6
)
ex  # tells me it is a tibble

mtcars  # Does not say what it is

```

#### 10.5.2    Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

```{r}
df <- data.frame(abc = 1, xyz = "a")
df
str(df)  # 1 observation of 2 variables
df$x  # I just get 'a';  this is due to partial matching of column name

df[, "xyz"] # this is as I would have expected

df[, c("abc", "xyz")] # as is this.  but a lot of typing

df.tib <- tibble(abc = 1, xyz = "a")
df.tib

df.tib$x # error message
df.tib[, "xyz"]  # returns in format that makes clear what I've asked for
df.tib[, c("abc", "xyz")] # ditto

```

#### 10.5.3   If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?

```{r}

str(mtcars)

mtcars.tib <- as_tibble(mtcars)

var <- "mpg"

mtcars.tib[[var]] # that worked

# mtcars %>% .$var  # doesn't work
# mtcars$var  # doesn't work

mtcars$mpg # does work

```

#### 10.5.4   Practice referring to non-syntactic names in the following data frame by:

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
# extract the variable called 1

annoying[1] # still a tibble
annoying[[1]]  # a vector
annoying$`1` # a vector

#Plotting a scatterplot of 1 vs 2.

ggplot(annoying, aes(`1`, `2`)) +
  geom_point()

# Creating a new column called 3 which is 2 divided by 1.

annoying.2 <- mutate(annoying,
                     `3` = `2`/`1`)
# OR
annoying[["3"]]  <- annoying[["2"]]/annoying[["1"]]

# Renaming the columns to one, two and three.

annoying.3 <- select(annoying.2, 1, 2,3) # that runs
annoying.3 <- rename(annoying.2, one = `1`, two = `2`, three = `3`) # that took some fiddling

```

#### 10.5.5   What does tibble::enframe() do? When might you use it?

```{r}

rainbow <- c("red", "orange", "yellow", "green", "blue", "indigo", "violet")

enframe(rainbow)  # makes a tibble.  automatically provides a rownumber column
 
#  enframe(letters[1:7], rainbow)  #error

# this only works with vectors of lists (to create a 2 column vdate frame)

```

Seems like a fast way to convert vector to dataframe. 


#### 10.5.6   What option controls how many additional column names are printed at the footer of a tibble?

This is controlled by how many columns you view the values for

Here are some setting options:

options(tibble.print_max = n, tibble.print_min = m): if more than m rows, print only n rows. 

Use options(dplyr.print_min = Inf) to always show all rows.

Use options(tibble.width = Inf) to always print all columns, regardless of the width of the screen.

```{r}
package?tibble
```

I find that: 
tibble.max_extra_cols
Number of extra columns printed in reduced form. Default: 100.