---
title: "Gene_expression_data_with_tidyr"
author: "Stacey Harmer"
date: "6/6/2017"
output: 
  html_document: 
    keep_md: yes
---

Working examples found at http://varianceexplained.org/r/tidy-genomics/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load the data
```{r}
library(readr)
original_data <- read_delim("http://varianceexplained.org/files/Brauer2008_DataSet1.tds", delim = "\t")

View(original_data)
summary(original_data)

original_data$NAME[1:3]
```

First, clean up the name information

```{r}
library(dplyr)
library(tidyr)

cleaned_data <- original_data %>%
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), sep = "\\|\\|")

head(cleaned_data)
View(cleaned_data)
```

Now get rid of extra whitespace

```{r}
cleaned_data <- original_data %>%
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), sep = "\\|\\|") %>%
  mutate_each(funs(trimws), name:systematic_name)
View(cleaned_data)

```

And now drop stuff we don't need
```{r}
cleaned_data <- original_data %>%
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), sep = "\\|\\|") %>%
  mutate_each(funs(trimws), name:systematic_name) %>%
  select(-number, -GID, -YORF, -GWEIGHT)
View(cleaned_data)
```

The data is not tidy; too much inforamtion in each row.
the gather() function will help here

```{r}

cleaned_data <- original_data %>%
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), sep = "\\|\\|") %>%
  mutate_each(funs(trimws), name:systematic_name) %>%
  select(-number, -GID, -YORF, -GWEIGHT) %>%
  gather(sample, expression, G0.05:U0.3)
View(cleaned_data)

```

That is really amazing!   Saved a bunch of work with melting.

Finally - need to separate the 2 types of information currently in the 'sample' column

```{r}
library(tidyverse)


cleaned_data <- original_data %>%
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), sep = "\\|\\|") %>%
  mutate_each(funs(trimws), name:systematic_name) %>%
  select(-number, -GID, -YORF, -GWEIGHT) %>%
  gather(sample, expression, G0.05:U0.3) %>%
  separate(sample, c("nutrient", "rate"), sep = 1, convert = TRUE)

View(cleaned_data)
```

Now for some plotting

```{r}
cleaned_data %>%
  filter(name == "LEU1") %>%
  ggplot(aes(rate, expression, color = nutrient)) +
  geom_line()
```

or more broadly

```{r}
cleaned_data %>%
  filter(BP == "leucine biosynthesis") %>%
  ggplot(aes(rate, expression, color = nutrient)) +
  geom_line() +
  facet_wrap(~name)

# fit lines to hte data
cleaned_data %>%
  filter(BP == "leucine biosynthesis") %>%
  ggplot(aes(rate, expression, color = nutrient)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~name)

```

Next up: http://varianceexplained.org/r/tidy-genomics-broom/

Modeling gene epxression 

And then http://varianceexplained.org/r/tidy-genomics-biobroom/
which 
use the limma and biobroom packages in combination to continue a tidy analysis, and consider when and how to use non-tidy data in an analysis.

Work through the above before going to:
https://simplystatistics.org/2016/02/17/non-tidy-data/