---
title: "DataTrans_part2_2017-05-10.Rmd"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Transformation (Ch 3 in book, Ch 5 online)

```{r}
library(nycflights13)
library(tidyverse)
head(flights)
```

#### 5.6  Group Summaries  
```{r}
by_dest <- group_by(flights, dest)
summary(by_dest)

summarize(by_dest, count = n())

delay <- summarize(by_dest, count = n(),
          dist = mean(distance, na.rm = T),
          delay = mean(arr_delay, na.rm = T))

delay <- filter(delay, count > 20, dest != "HNL")
```

Pipes!
```{r}
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

```

batting data
```{r}
batting <- as_tibble(Lahman::Batting)

batters <- batting %>%
  group_by(playerID) %>%
    summarize(
      ba = sum(H, na.rm = T) / sum(AB, na.rm = T),
      ab = sum(AB, na.rm =T)
    )

```

weird!  When I summarize(batters), I get
"data frame with 0 columns and 0 rows"
but whtn I type 'batters', I see a dataframe

```{r}
batters %>%
  filter(ab > 100) %>%
    ggplot(mapping = aes(x = ab, y = ba)) +
  geom_point() +
  geom_smooth(se = F)

```
subsetting

```{r}
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(
    avg_delay1 = mean(arr_delay), 
    avg_delay2 = mean(arr_delay[arr_delay > 0])
  )
```
Measures of spread

```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarize(distance_sd = sd(distance)) %>%
  arrange(desc(distance_sd))
```

Rank
```{r}
not_cancelled %>%
  group_by(year, month, day) %>%
  mutate(r = min_rank(desc(dep_time))) %>%
  filter(r %in% range(r))
```
Count
```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarize(carriers = n_distinct(carrier)) %>%
  arrange(desc(carriers))
```

Proportion: note that when used with numeric functions, mean can tell you proportion
```{r}

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(hour_perc = mean(arr_delay > 60))
```
grouping by multiple variables
```{r}
daily <- group_by(flights, year, month, day)
head(daily, n =2)

per_day <- summarize(daily, flights = n())
head(per_day, n = 2)

per_month <- summarize(per_day, flights = sum(flights))
head(per_month, n =2)  

per_year <- summarize(per_month, flights = sum(flights))
head(per_year, n = 2)
```
ungroup

```{r}
daily %>%
  summarize(flights = n()) #summarized by flights per day

daily %>%
  ungroup() %>%
  summarize(flights = n()) # summary of all flights all year

```

#### 5.6.7  Exercises

##### 5.6.7.1  Think of at least 5 ways to assess typical delay characteristics of a group of flights.  

Which is more important?  Arrival delay or departure delay?  
To me, arrival delay. This can mean you'll miss your next connection.

How to find flights that are:

 15 min early 50% of time, and 15 min late 50% of time:
```{r}
delay.15 <- not_cancelled %>%
  group_by(flight) %>%
  summarize(
    late_15 = mean(arr_delay == 15),
    early_15 = mean(arr_delay == -15)
    ) %>%
  filter(late_15 == .5, early_15 == .5)
# no such flights exist
 
#did it work?  Try only one filter
delay.15 <- not_cancelled %>%
  group_by(flight) %>%
  summarize(
    late_15 = mean(arr_delay == 15),
    early_15 = mean(arr_delay == -15)
    ) %>%
  filter(late_15 == .5)
# Yes, 3 flights found
```

A flight is always 10 minutes late.  
I could easily modify the above.  But what would be a different way?
```{r}
#  str(not_cancelled) tells me there are 327,346 rows in dataframe(observations)
#  not_cancelled %>% count(flight) tells me there are 3,835 flights in dataset

delay.10 <- not_cancelled %>%
  group_by(flight) %>%  # 3,835 total flights when grouped this way
  filter(arr_delay == 10) %>% # sort by flight num to be sure it worked
  arrange(flight)
# although of course it doesn't tell us that ALL versions of that flight are delayed 10 min
# str(delay.10) tells me there are 3,373 observations.  
# I think that I have retrieved the rows in which arr_delay is 10 min, and that this happened at least once in all cases

delay.10 <- not_cancelled %>%
  group_by(flight) %>%  
  summarize(
    perc_late_10 = mean(arr_delay == 10)
    )

delay.10 <- not_cancelled %>%
  group_by(flight) %>%  
  summarize(
    perc_late_10 = mean(arr_delay == 10)) %>%
    filter(perc_late_10 == 1)  # there are 5 such flights

# can I move helpful information into output?
# I couldn't figure out how

```

A flight is 30 minutes early 50% of the time, and 30 min late 50% of the time

```{r}
delay.30 <- not_cancelled %>%
  group_by(flight) %>%
  summarize(
    late_30 = sum(arr_delay == 30),
    early_30 = sum(arr_delay == -30),
    count = n()
    ) %>%  
  filter(late_30/count == .5, early_30/count == .5) 
# nothing came up with this filter

delay.30 <- not_cancelled %>%
  group_by(flight) %>%
  summarize(
    late_30 = sum(arr_delay == 30),
    early_30 = sum(arr_delay == -30),
    count = n()
    ) %>%   #now I need to filter so sum/total = .5
  filter(late_30/count == .5 )
# two flights are late 30 min half the time; only 2 flights, though

```
99% of the time a flight is on time.  1% of the time it is 2 hr (120 min) late

```{r}
timely.99 <- not_cancelled %>%
  group_by(flight) %>%
  filter(arr_delay == 0 | arr_delay == 120) %>%
  summarize(
    on_time_99 = mean(arr_delay == 0),
    count = n() 
  ) %>%
filter(on_time_99/count == 0.99)

# no flights fit this criterion
```


##### 5.6.2  Think of a new way to accomplish each of the following
(without using count)
```{r}
not_cancelled %>%
  count(dest) # this sums up total # flights to each airport

not_cancelled %>%
  group_by(dest) %>%
  summarize(
    num_flights = n()) # that works

not_cancelled %>%
  group_by(dest) %>%
  summarize(
    num_flights = sum(!is.na(flight))) #also works

```


```{r}
not_cancelled %>%
  count(tailnum, wt = distance) # this sums total # miles plane flew

not_cancelled %>%
  group_by(tailnum) %>%
  summarize(
    distance_plane = sum(distance))  # works

```

##### 5.6.3  what is wrong with defining cancelled flights as those that either had an NA for departure delay or an NA for arrival delay?

Seems like a pretty good definition to me.  Let's look at examples
```{r}

no_dep_delay <- filter(flights, is.na(dep_delay), !is.na(arr_delay))
# nothing matches this; all flights with NA for dep_delay also have NA for arr_delay.  Makes sense.

no_arr_delay <- filter(flights, !is.na(dep_delay), is.na(arr_delay))
#  About 1200 flights have NA for arrival delay, but do report a dep_delay.

View(no_arr_delay)
summary(no_arr_delay)
```
Some have an arrival time, some do not. (all have at least a scheduled arrival time.) None have an air time.
Not clear to me what happened.  

##### 5.6.4  Look at the number of cancelled flights per day.  Is there a pattern?  Is the proportion related to average delay?
I will use the flights with NA for arr_delay as cancelled flights

```{r}

flights %>%
  group_by(year, month, day) %>%
  summarize(
    perc_cancel = mean(is.na(arr_delay)),
    avg_delay = mean(dep_delay, na.rm = T), 
    n = n()
    ) %>%   #  pipe this to ggplot
ggplot(mapping = aes(x = avg_delay, y = perc_cancel)) +
  geom_point(alpha = 0.3)
```
As you might expect, more cencellations on days with longer average delays.

##### 5.6.5  Which carrier has the worst delays?
```{r}
# simple approach

rank.delays <- not_cancelled %>%
  group_by(carrier) %>%
  summarize(
    avg_dep_delay = mean(dep_delay),
    count = n()
  ) %>%
  arrange(desc(avg_dep_delay))

airlines # 9E is the worst, it is Endeavor Air.  Then ExpressJet, then Mesa
 
```
Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))

```{r}

rank.delays.2 <- not_cancelled %>%
  group_by(carrier,dest) %>%
  summarize(
    avg_dep_delay = mean(dep_delay),
    count = n()
  ) %>%
  arrange(desc(dest, avg_dep_delay))
```
How would I plot that?  Maybe dest by delay, facet by carrier?
```{r}
not_cancelled %>%
  group_by(carrier,dest) %>%
  summarize(
    avg_dep_delay = mean(dep_delay),
    count = n()
  ) %>%
  ggplot(mapping = aes(x = carrier, y = avg_dep_delay)) +
  geom_col() +
  facet_wrap( ~ dest)
# that is too busy to really be useful.  
```

##### 5.6.6  What does the sort() argument to count() do?

```{r}
# the help page tells me sort sorts output in descending order
# might want to do this to examine # of flights to each destination

sort_flights <- not_cancelled %>%
  group_by(carrier,dest) %>%
  count(flight, sort = T)
# something seems wrong about the output, but I don't really wnat to figure it out now

```

#### 5.7  Group Mutates
window functions are used in conjuncton with mutate and filter

##### 5.7.1  Refer back to the lists of useful mutate and filtering functions. Describe how each operation changes when you combine it with grouping.

Well, that sounds tedious.  Useful mutate functions include arithmetic, log, offsets,ranking.  These all naturally work with groups.  Logical comparisons end up giving you T/F results that can be added.

Useful filter functions include comparisons, logical operators.  Nice to use with groups to get overview of traits.

##### 5.7.2  Which plane (tailnum) has worst on-time record?
```{r}
# maybe the easiest thing is to compute fraction of arr_delay == 0.

ranked_delay <- not_cancelled %>%
  group_by(tailnum) %>%
  mutate(
    on_time = mean(arr_delay == 0),  # planes never on time
    avg_delay = mean(arr_delay), # then average delay
    count = n()
  ) %>%
  select(year:day, carrier, tailnum, on_time, avg_delay, count) %>%
  arrange(on_time, avg_delay)

head(ranked_delay)  # looks good
```

##### 5.7.3  What time of day should you fly to avoid delays?
```{r}

str(not_cancelled) # hour should be useful

by_time <- not_cancelled %>%
  group_by(hour) %>%
  summarize(
    avg_delay = mean(arr_delay, na.rm = T),
    count = n()
  ) %>%
  ggplot(mapping = aes(x = hour, y = avg_delay)) +
  geom_point()

by_time
# makes sense.
```

##### 5.7.4  Compute total number of minutes of delay for each dest.  For each flight compute proportion of the total delay for its destination.

Compute total number of minutes of delay for each dest.
```{r}
not_cancelled %>%
  count(dest, wt = arr_delay)

```
For each flight compute the proportion of the total delay for its destination.
(I think I am supposed to figure out what fraction of the above total is due to each flight)
```{r}
not_cancelled %>%
  group_by(dest) %>%
  mutate(
    flight_delay = sum(arr_delay),
    perc_tot_delay = arr_delay/sum(arr_delay)
  ) %>%
  select(year:day, arr_delay, carrier, flight, dest, flight_delay, perc_tot_delay) %>%
  arrange(desc(perc_tot_delay))
```

##### 5.7.5  Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using lag() explore how the delay of a flight is related to the delay of the immediately preceding flight.

```{r}
# I think the most direct thing to do would be to sort by time and then graph time and lag time for dep_delay

not_cancelled %>%
  mutate(
    previous_dep_delay = dep_delay,
    next_dep_delay = lag(previous_dep_delay)
  ) %>%
  ggplot(mapping = aes(x = previous_dep_delay, y = next_dep_delay)) + 
  geom_point(, aes = 0.2) +
  geom_smooth()
# not as direct a relationship as you might expect

```

##### 5.7.6  Look at each destination. Can you find flights that are suspiciously fast? (i.e. flights that represent a potential data entry error). Compute the air time a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?

Can you find flights that are suspiciously fast? Compute the air time a flight relative to the shortest flight to that destination
```{r}
times <- not_cancelled %>%
  group_by(dest) %>%
  mutate(fastest = min(air_time)) %>%
  arrange(dest, desc(air_time))

ggplot(times, aes(x = air_time, y = fastest)) +
  geom_point(alpha = .2) + 
  facet_grid(.~dest)
# too many points to be useful

# so let's sort instead

times <- not_cancelled %>%
  group_by(dest) %>%
  mutate(flight_time_mean = mean(air_time),
         fold_flight_mean = air_time / flight_time_mean
         ) %>%
  arrange(fold_flight_mean) %>%
  select(year:day, flight_time_mean, air_time, fold_flight_mean)

head(times)

```

##### 5.7.7  Find all destinations flown by at least 2 carriers.  then rank the carriers (?)
```{r}
not_cancelled %>%
  select(year:day, carrier, flight, tailnum, dest, arr_delay) %>%
  group_by(dest, carrier) %>%
  count(carrier) 
# This tells me how many flights each carrier had to the given destinations

not_cancelled %>%
  select(year:day, carrier, flight, tailnum, dest, arr_delay) %>%
  group_by(dest, carrier) %>%
  count(carrier) %>%
  arrange(dest, desc(n))
# and this sorts by # of flights per carrier to each destination

```

##### 5.7.8  For each plane, count the number of flights before the first delay of greater than 1 hour.
Not quite sure what to do.  I think use tailnum as sorting criterion

```{r}

num_b4_delay <- not_cancelled %>%
  group_by(tailnum) %>%
  filter(arr_delay < 60)

head(num_b4_delay)
summary(num_b4_delay) # not quite right.  Want to count flights before dleay > 60.  Stuck!

```

#### Tips

1.  new shortcut : Ctrl + Shift + 1/2/3/4 to turn one of the panes to full-screen - nice!

2.  Switch tabs w/o muscle cramps: switch between open tabs w/ Ctrl+Tab. (go the other direction with Ctrl + Tab + Shift)

3.  Jump quickly between the source and console via keyboard: Ctrl+1 to focus source, Ctrl+2 to focus console.  

4. Ctrl + shift + M gives the pipe ( %>% )


