---
title: "R Notebook"
output: 
  html_document: 
    keep_md: yes
---
Section 21: part 4 - end

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 21.4
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

output <- vector("double", length(df))
for (i in seq_along(df)) {
  output[[i]] <- mean(df[[i]])
}
output

```

make that a function

```{r}
col_mean <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- mean(df[[i]])
  }
  output
}

col_mean(df)

```

but, we can pass one function to another!  
```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
col_summary(df, median)
col_summary(df, sum)

```

##21.4.1 Exercises

1.  Read the help file for apply().  In the 2d case, what two for loops does it generate?

I have no idea what the 2d case means. 

2.  Adapt col_summary() so that it only applies to numeric columns 
```{r}
#  start with an is_numeric() function that returns a logical vector that has a TRUE corresponding to each numeric column.

col_summary2 <- function(df, fun) {
  # test whether each colum is numeric
  numeric_cols <- vector("logical", length(df))
  for (i in seq_along(df)) {
    numeric_cols[[i]] <- is.numeric(df[[i]])
  }
  # indexes of numeric columns
  idxs <- seq_along(df)[numeric_cols]
  # number of numeric columns
  n <- sum(numeric_cols)
  out <- vector("double", n)
  for (i in idxs) {
    out[i] <- fun(df[[i]])
  }
  out
}

# whoa, that was complicated

col_summary_rev <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

```

# 21.5  The map functions

```{r}
map_dbl(df, mean)

map_dbl(df, median)

```

An even more transparent method:
```{r}

df %>% map_dbl(mean)
```

## 21.5.1  Shortcuts

```{r}
models <- mtcars %>% 
  split(.$cyl) %>%  # that is handy! 
  map(function(df) lm(mpg ~ wt, data = df))

# or

models <- mtcars %>% 
  split(.$cyl) %>% 
  map(~lm(mpg ~ wt, data = .))
```

 now extract summary
 
```{r}
models %>% 
  map(summary) %>% 
  map_dbl(~.$r.squared)

# or more concisely
models %>% 
  map(summary) %>% 
  map_dbl("r.squared")

```

#  21.5.2  Base R

```{r}
x1 <- list(
  c(0.27, 0.37, 0.57, 0.91, 0.20),
  c(0.90, 0.94, 0.66, 0.63, 0.06), 
  c(0.21, 0.18, 0.69, 0.38, 0.77)
)
x2 <- list(
  c(0.50, 0.72, 0.99, 0.38, 0.78), 
  c(0.93, 0.21, 0.65, 0.13, 0.27), 
  c(0.39, 0.01, 0.38, 0.87, 0.34)
)

threshold <- function(x, cutoff = 0.8) x[x > cutoff]
x1 %>% sapply(threshold) %>% str()
x2 %>% sapply(threshold) %>% str()
```

## 21.5.3  MAP function exercises

1.  Write code that uses one of the map functions to:

1. Compute the mean of every column in mtcars.

```{r}
head(mtcars)

map_dbl(mtcars, mean) # easy!
```

2. Determine the type of each column in nycflights13::flights.
```{r}
head(nycflights13::flights)
map_chr(nycflights13::flights, typeof)
```
doesn't get the datetime right for last column
```{r}
map(nycflights13::flights, class)
# this gives list of differnet types
```

3. Compute the number of unique values in each column of iris.

```{r, error = TRUE}
head(iris)
map_dbl(iris, length(unique())) # didn't like that 

map_dbl(iris, function(df) length(unique(df)), data = df) # didn't like that
# I think length tried to cope wiht entire df, not column
map_dbl(iris, function(df) length(unique()), data = df) # same problem
map_dbl(iris, function(df) length(unique(df))) # this works

map_dbl(iris, ~ length(unique(.))) # this works

```

4. Generate 10 random normals for each of   μ = -10, 0, 10, and 100

```{r, error = TRUE}

x <- 0
rnorm(10, mean = x)

x <- c(-10,0,10,100)

map_dbl(n = 10, rnorm, mean = x) # nope

map(n = 10, rnorm, mean = x) # nope

map(x, rnorm, n = 10) # order matters here b/c map will split up the first element (either by column or by element)
```

2. How can you create a single vector that for each column in a data frame indicates whether or not it’s a factor?

```{r}
map_lgl(df, is.factor)

```

3. What happens when you use the map functions on vectors that aren’t lists? What does map(1:5, runif) do? Why?

```{r}
?runif
runif(10)
map(1:5, runif)

```

Generated a list of 5 elements. 

4. What does map(-2:2, rnorm, n = 5) do? Why? 
```{r}

 map(-2:2, rnorm, n = 5)
```

Generates list with 5 elements, each with 5 components.  rnorm

What does map_dbl(-2:2, rnorm, n = 5) do? Why?
```{r, error = T}
map_dbl(-2:2, rnorm, n = 5)

```
This fails to run.  presumably because it can't handle a list
```{r}
rnorm(0, n = 5)
```

5. Rewrite map(x, function(df) lm(mpg ~ wt, data = df)) to eliminate the anonymous function

```{r, error = TRUE}

map(x, function(df) lm(mpg ~ wt, data = df))

map(x, ~lm(mpg ~ wt, data = .))

list(mtcars)

map(list(mtcars), ~lm(mpg ~ wt, data = .)) # more complete
```

#21.6   Dealing with failure

```{r}
x <- list(1, 10, "a")
y <- x %>% map(safely(log))
str(y)

transpose(y)
y <- y %>% transpose()
str(y)

is_ok <- y$error %>% map_lgl(is_null)
x[!is_ok]
#> [[1]]
#> [1] "a"
y$result[is_ok] %>% flatten_dbl()


```

```{r}
list(1, -1)
```

#21.7  Mapping over multiple arguments

map2()

Note that the arguments that vary for each call come before the function; arguments that are the same for every call come after.

##21.7.1  Invoking differnet functions

```{r}

f <- c("runif", "rnorm", "rpois")
param <- list(
  list(min = -1, max = 1), 
  list(sd = 5), 
  list(lambda = 10)
)


sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)
sim[[2]]
sim[[1]]

sim %>% 
  mutate(sim = invoke_map(f, params, n = 10))

```

#21.8  Walk

```{r}
x <- list(1, "a", 3)

x %>% 
  walk(print)

?pwalk
```

#21.9  Other patterns of for loops

##21.9.1  Predicate

keep and discard

```{r}
iris %>% 
  keep(is.factor) %>% 
  str()


iris %>% 
  discard(is.factor) %>% 
  str()
```

some and every

```{r}
x <- list(1:5, letters, list(10))

x %>% 
  some(is_character)
#> [1] TRUE

x %>% 
  every(is_vector)
```

detect - finds first element

```{r}
x <- sample(10)
x

x %>% 
  detect(~ . > 5)

x %>% 
  detect_index(~ . > 5)
```

head_while

```{r}
x
x %>% 
  head_while(~ . > 5)

x %>% 
  tail_while(~ . > 5)
```

## 21.9.2  reduce and accumulate

```{r}
dfs <- list(
  age = tibble(name = "John", age = 30),
  sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
  trt = tibble(name = "Mary", treatment = "A")
)

1:3 %>% reduce(`+`)
reduce(1:3, `+`)
reduce(1:3, sum)

dfs %>% reduce(full_join)
```

Or

```{r}
vs <- list(
  c(1, 3, 5, 6, 10),
  c(1, 2, 3, 7, 8, 10),
  c(1, 2, 3, 4, 8, 9, 10)
)

vs %>% reduce(intersect)

reduce(vs, intersect)
```

## 21.9.3  Exercises

1. Implement your own version of every() using a for loop. Compare it with purrr::every(). What does purrr’s version do that your version doesn’t?

```{r}
x <- list(1:5, letters, list(10))

x %>% 
  every(is_vector)
```

I want the loop to examine every vector in my list and tell me if each meets criterion

```{r}

output <- vector("logical", ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output

my_every <- function(df, fun) {
  out <- vector("logical", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

x <- list(1:5, letters, list(10))

x %>% 
  every(is_vector)

my_every(x, is_vector)

```

2. Create an enhanced col_sum() that applies a summary function to every numeric column in a data frame.

```{r, error = TRUE}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

col_summary(iris, median) # error

col_summary2 <- function(df, fun) {
  temp <- keep(df, is.numeric)
  out <- vector("double", length(temp))
  for (i in seq_along(temp)) {
    out[i] <- fun(temp[[i]])
  }
  out
}

col_summary2(iris, median)

# or more simply

col_summary3 <- function(df, f, ...) {
  map(keep(df, is.numeric), f, ...)
}

col_summary3(iris, median)

```

3. A possible base R equivalent of col_sum() is:

```{r}

col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}

```

But problems may arise:
```{r}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)

# OK
col_sum3(df, mean) doesn't return anything for non-numeric column
col_summary2(df, mean) #  my version is same

# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_summary2(df[1:2], mean) 

col_sum3(df[1], mean)
col_sum3(df[0], mean)

```

