---
title: "Ch 12 - tidy data, part 1"
author: "Stacey Harmer"
date: "6/11/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Sections 12.1 - 12.4

#12 Tidy data

## 12.2.1  Tidy data exercises

### 12.2.1.1 Describe how the variables and observations are organised in each of the sample tables.

Table 1: each column is a variable, and every row is an observation
Table 2:  one of the columns ('type') is a factor, while the others are variables
Table 3:  one of hte colums ('rate') has two observations in it
Table 4:  two tibbles.  'year' is a factor now, and is spread over 2 columns in each tibble.

### 12.2.1.2 Compute the rate for table2, and table4a + table4b. You will need to perform four operations:
1.  extract # TB cases per country per year, and population # as well
2.  divide cases by population, then multiply by 10000
3.  Store in correct locations

```{r}
library(tidyverse)
head(table2)

#table2 %>%
 # group_by(country, year) %>%
  #mutate(rate = filter(type == "cases")/filter(type == "population") )

kludge <- cbind(table2 %>%
  group_by(country, year) %>%
   filter(type == "cases"),
  table2 %>%
  group_by(country, year) %>%
   filter(type == "population"))

kludge %>%
  transform(country, year, 
            rate = kludge[, 4]/kludge[, 8] * 10000)

# Ugly, but it works

# better way: make a new tibble from extracted vectors
table2_cases <-  filter(table2, type == "cases")[, c(1:2,4)]
table2_pop <- filter(table2, type == "population")[, 4]

table2_new <- bind_cols(table2_cases, table2_pop)
colnames(table2_new)[3:4] <- c("cases", "population") 

table2_new %>%
  mutate(rates = cases/population * 10000)

```
That was a drag.  Now do the same for Table 4

```{r}
table4a

table4b

(rates <- tibble(rate.1999 = table4a$'1999'/table4b$'1999' *10000,
                rate.2000 = table4a$'2000'/table4b$'2000' * 10000))

(table.4a.rates <- bind_cols(table4a[1], rates))

#this worked; but more elegant answer on solutions page

tibble(country = table4a[["country"]],
       '1999' = table4a[['1999']]/ table4b[['1999']] * 10000,
       '2000' = table4a[['2000']]/ table4b[['2000']] * 10000
       )
```

### 12.2.1.3  Recreate the plot showing change in cases over time using table2 instead of table1. What do you need to do first?

```{r}
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

table2

```

To plot this, I first need to filter out the population data

```{r}

table2 %>%
  filter(type == "cases") %>%
ggplot(aes(year, count)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```
That was pretty easy.

## 12.3  Spreading and gathering
### 12.3.1  Gathering
makes wide tables longer and narrower

when some column names are values of a variable, not a variable name

```{r}

tidy4a <- table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")

tidy4b <- table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")

# left join is useful here
left_join(tidy4a, tidy4b) # smart enough not to duplicate country and year


```

### 12.3.2  Spreading
makes long tables shorter and wider

what is the key column? the one with variable names

```{r}
spread(table2, key = type, value = count)

```

### 12.3.3  Exercises

#### 12.3.3.1  Why are gather() and spread() not perfectly symmetrical?
Consider:

```{r}

stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)

```

the spread function made a wider tibble; colnames now half and 2015 and 2016; 
these are all numbers (dbl)  (same as original tibble)

Running gather on that - now year is a character.  this seems weird.
Gather converted the variable name to a character

Try again but with convert = T
```{r}
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`, convert = T)

```
If I do this, year is an integer.
Help file says:  "if convert = T, will automatically run type.convert on key column"
type.convert will convert a character vector to logical, numeric, complex, or factor

Not fully symmetrical because column type information is lost!


#### 12.3.3.2  Why does this code fail?

```{r}
#table4a %>% 
 # gather(1999, 2000, key = "year", value = "cases")

#> Error in combine_vars(vars, ind_list): Position must be between 0 and n
```

Lack of tick marks, I think.

```{r}

table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")

```

Yep.

```

#### 12.3.3.3  Why does spreading this tibble fail? How could you add a new column to fix the problem?

```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
```

I could add a column 'observation' c(1,1,2,1,1)

```{r}
#  people2 <- right_join(people, c(1,1,2,1,1))
# this doesn't work
# cbind?
people2 <- cbind(people, c(1,1,2,1,1))
# that works, but messy
colnames(people2)[4] <- "obs"

people2 <- cbind(people, obs = c(1,1,2,1,1))
# this also works, and is more concise

spread(people2, key, value)

```

Or, I think if I got rid of one row that would do it also

```{r}
 # ( although data is lost here)
people3 <- people[c(1:2, 4:5),]

spread(people3, key = key, value = value )

```


#### 12.3.3.4  Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
# the variables are pregnant, and sex
# so I need to gather it

gather(preg, male, female, key = "sex", value = "count")

# the help page points out that logical vectorw sould be easier to work with

gather(preg, male, female, key = "sex", value = "count") %>%
  mutate(pregnant = pregnant == "yes",
         female = sex == "female") %>%
  select(-sex)

```

## 12.4  Separating and uniting

### 12.4.1  Separating

```{r}
library(tidyverse)
table3
table3 %>% 
  separate(rate, into = c("cases", "population"))

```
Default is to leave column type as is (in this case, characters)

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"),convert = T)

table3 %>% 
  separate(year, into = c("century", "year"), sep = 2, convert = T)

table3 %>% 
  separate(year, into = c("century", "year"), sep = 2) %>%
  separate(rate, into = c("cases", "population"), convert = T)
```

Nice!

```{r}
table5
table5 %>% 
  unite(new, century, year, sep = "")
```


### 12.4.3  Exercises

#### 12.4.3.1 What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets.

```{r}

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"))

# default value for extra is warn; I got a warning telling me that one of my rows had
# more values than the others
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "drop")
# same output, no warning

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")
# no values dropped, instead the 'extra' value left in the last cell of 2nd row

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))

# fill default value is 'warn'.  told me an NA was inserted

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "right")

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")

# sets where the NA gets inserted.  
```

#### 12.4.3.2 Both unite() and separate() have a remove argument. What does it do? Why would you set it to FALSE?

The help file tells me that default is remove; this removes the input columns from output df.  Makes sense; usually don't want same information twice. 

Might imagine wanting it to be false if: you want to check the conversion is accurate.  Or if you for some reason both want full value and abbreviated value.

```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2, remove = F)

```

#### 12.4.3.3 Compare and contrast separate() and extract(). Why are there three variations of separation (by position, by separator, and with groups), but only one unite?

Separate with groups?  not sure what that means.  And what is the extract() function?  Does he mean []?
We were introduced to hte subsetting function in section 10.3.2


```{r}

df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

df
# remember: To use these in a pipe, you’ll need to use the special placeholder .:

df %>% .$x

```

