---
title: "2017-09-06_iteration_part1.Rmd"
author: "Stacey Harmer"
date: "9/6/2017"
output: 
  html_document: 
    keep_md: yes
---

Ch 21 (for loops/iteration) through 21.4

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Imperative programming = for loops, and while loops
functional programming = each common for loop gets its own function

## 21.2  For loops

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df

output <- vector("double", ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output

## WHY does seq_along count columns??  

output <- vector()  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output
```

## 21.2.1  For loop exercises

### 1. Write for loops to:

1. Compute the mean of every column in mtcars.
```{r}
head(mtcars)
summary(mtcars)

mtcars.means <- vector("double", ncol(mtcars))  
     for (i in seq_along(mtcars)) {            
     mtcars.means[[i]] <- mean(mtcars[[i]])      
   }
mtcars.means

output <- vector("double", ncol(mtcars))
names(output) <- names(mtcars) # nice touch

```

2. Determine the type of each column in nycflights13::flights.

```{r}
head(nycflights13::flights)

flight.types <- vector("character", ncol(nycflights13::flights))
for (i in seq_along(nycflights13::flights)) {
  flight.types[[i]] <- typeof(nycflights13::flights[[i]])
}

flight.types
# I see that the last column was 'dttm' in the tibble but is 'double' when using typeof()

flight.types <- vector("list", ncol(nycflights13::flights))
for (i in seq_along(nycflights13::flights)) {
  flight.types[[i]] <- class(nycflights13::flights[[i]])
}

flight.types # this returned a date time for the last value

```

3. Compute the number of unique values in each column of iris.
```{r}
head(iris)
summary(iris)

# I think I need to use unique and length.  Try it:

length(unique(iris$Species)) # yep

unique.iris <- vector("double", ncol(iris))
for (i in seq_along(iris)) {
  unique.iris[[i]] <- length(unique(iris[[i]]))
}
unique.iris

```

4.  Generate 10 random normals for each of  
μ = -10, 0, 10, and 100

```{r, error = TRUE}
mu <- c(-10, 0, 10, 100)
rnorm(10, -10)
# I think I'd like a dataframe rather than a vector, but will work with vector here

# should have made a matrix or a tibble, not a vector

mu.ran.norm <- vector("double", 40)
for (i in mu) {
  mu.ran.norm[[i]] <- rnorm(10, mu[[i]])
}

rnorm(10, mu[[1]]) # this works
# try with a list instead of a vector

mu.ran.norm <- vector("list", length(mu))
for (i in seq_along(mu)) {
  mu.ran.norm[[i]] <- rnorm(10, mu[[i]])
}

mu.ran.norm

```

### 2. Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

The below is a paste function
```{r}

out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
library(stringr)
str_c(letters, sep = "")
length(str_c(letters, sep = ""))
str_c(letters, collapse = "")

```


A roundabout way to calculate the std dev

```{r}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd
sd <- sqrt(sd / (length(x) - 1))

sd(x) # easy!
```
 
 

```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1] # i guess this is just to give a starting valule for out[i]

for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out

# cumulative sum

cumsum(x)

```

### 3. Combine your function writing and for loop skills:

Write a for loop that prints() the lyrics to the children’s song “Alice the camel”.

x <- c(5, 4, 3, 2, 1, 0)
x <- c("five","four","three", "two", "one", "no")

Alice the camel has five humps.
for (i in seq_along(x)) {
print("Alice the camel has")


### 4. It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:

```{r, error = TRUE}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
x
lengths(x[[995]])
```
Test how this affects performance vs pre-allocating vector output

```{r}
x <- rnorm(10000)
output <- vector("integer", 0)
ptm <- proc.time()
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
proc.time() - ptm

#   user  system elapsed 
#  0.161   0.072   0.237 

#now preallocate
x <- rnorm(10000)
output <- vector("integer",  length(x))

ptm <- proc.time()
for (i in seq_along(x)) {
  output[[i]] <-  lengths(x[[i]])
}
proc.time() - ptm

#    user  system elapsed 
#   0.384   0.158   0.546

```

Julin helped me fix it - see above.

## 21.3  For loop variations
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df
x <- c(0, 5, 15,22, 60)
range(df$a)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
df

# or, try the below
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df

for (i in seq_along(df)) {
  df[[i]] <- rescale01(df[[i]])
}
df
```

note:
Typically you’ll be modifying a list or data frame with this sort of loop, so remember to use [[, not [. You might have spotted that I used [[ in all my for loops: I think it’s better to use [[ even for atomic vectors because it makes it clear that I want to work with a single element.

#### 21.3.4 while loop

```{r}
sample(c("T", "H"), 1)
sample(c("T", "H"), 10, replace = T)

flip <- function() sample(c("T", "H"), 1)

flips <- 0
nheads <- 0
flip()

while (nheads < 3) {
  if (flip() == "H") {
    nheads <- nheads + 1
  } else {
    nheads <- 0
  }
  flips <- flips + 1
}
flips
```


#### 21.3.5.  Exercises

##### 1. Imagine you have a directory full of CSV files that you want to read in. 
You have their paths in a vector, files <- dir("data/", pattern = "\\.csv$", full.names = TRUE), and now want to read each one with read_csv(). Write the for loop that will load them into a single data frame.


```{r, error= T}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
# list of names of files

contents <- vector("list", length(files))

for (i in seq_along(files)){
  contents[i] <- read_csv(files[[i]])
}

contents <- bind_row(contents)

```

#### 2. What happens if you use for (nm in names(x)) and x has no names? What if only some of the elements are named? What if the names are not unique?

```{r}
x <- c(A = 1, B = 2, C = 3, C = 10)
names(x)

for (ID in names(x)){
  print(ID)
}

x <- c(1, 5, 10, 15)
names(x) # NULL

for (ID in names(x)){
  print(ID)
}
 # nothing, not even an error

x <- c(A = 1, 5, 2, C = 3, C = 10)
names(x)

for (ID in names(x)){
  print(ID)
}
# at least here you get an indication that some have no names

```

#### 3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, show_mean(iris) would print:

show_mean(iris)
 Sepal.Length: 5.84
 Sepal.Width:  3.06
 Petal.Length: 3.76
 Petal.Width:  1.20
 
 
```{r}

head(iris)
names(iris) # first 4 columns are numeric

iris.mean <- function(df) {
  for(ID in names(df)){
    if(is.numeric(df[[ID]])) {
      cat(ID, mean(df[[ID]]))
    }
  }
}

iris.mean(iris)
# not so pretty, but so what?

```

#### 4. What does this code do? How does it work?
```{r, error = TRUE}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)

# OK, this makes a list, disp = x* 0.016 and am now a factor.

for (var in names(trans)) { # that is, disp and am
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
head(mtcars)

# columns are titled disp (Displacement) and am (automatic = 0, manual = 1)

trans(mtcars[[disp]])(mtcars[[disp]]) # fail

trans[["am"]](mtcars[["am"]]) # ok, that makes sense

```
 
##  21.4  Functionals and for loops

```{r}

df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df

# compute mean for every column
output <- vector("double", length(df))
for (i in seq_along(df)) {
  output[[i]] <- mean(df[[i]])
}
output

# make this a function

col_mean <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- mean(df[[i]])
  }
  output
}

col_mean(df)
```

but how to generalize?

```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

col_summary(df, median)
col_summary(df, mean)
col_summary(df, sd)
```

