---
title: "DataVis_part1_2017-04-26.Rmd"
author: "Stacey Harmer"
date: "4/25/2017"
output: 
  html_document: 
    keep_md: yes
---
for the below chunk, include = FALSE means code in this chunk not included knitted file (include = F means neither code nor results of running code not shown)
But the line below sets defaults for page, and that is to show the code in the knitted file (echo = T shows code; echo = F means code not shown but the output of the code is shown)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Section 3.2: First steps
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

```

#### 3.2.4 Exercises
##### 1. run ggplot(data = mpg)
```{r}
ggplot(data = mpg)
```
I get a blank plot, with no data populating it.

##### 2.  How many rows and columns are in mtcars?
```{r}
dim(mtcars)
```
There are 32 rows and 11 columns in this dataframe

##### 3.  what does drv describe?
```{r}
?mpg
```
drv describes the transmission: f = front-wheel drive, r = rear wheel drive, 4 = 4wd

##### 4.  make scatterplot of hwy vs cyl
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cyl, y = hwy))
```

##### 5.  make scatterplot of class vs drv.  why not useful?
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = class, y = drv))
```

These data are overplotted.  Can't tell how many data points per position.

###  3.3: Aesthetic mappings
Can use levels (aesthetic properties) of data to provide more info in graph.

**NOTE** when shaped used as aesthetic, default is only 6 shapes per plot

_Figure 3.1_ shows the various shapes of points

#### 3.3.1 Exercises

##### 1.  What is wrong with below code?
```{r}
ggplot(data = mpg ) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

If points are supposed to be blue, code should be
```{r}
ggplot(data = mpg ) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

##### 2.  Which variables in mpg are categorical and which are continous?
```{r}
summary(mpg)
```
Categorical: manufacturer, model, trans, drv, fl, class

Continuous:  displ, year, cyl, cty, hwy

##### 3.  Map a continuous variable to color, size, and shape.

`ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ, size = year, shape = cyl))`
  
I ran the code; got an error message: continous variable can't be mapped to shape.  I omitted the code from this file to avoid knitting problems.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ, size = year))
```

But I can map continuous variables to displ and to size. 
Try again, but this time choose variables with larger range

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ, size = cty))
```

Dealt with that fine; just binned data.

##### 4.  Map same variable to multiple aesthetics
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = cty, color = drv, shape = drv))
```
This works.

##### 5.  Explore the stroke aesthetic.
```{r}
ggplot(data=mpg) +
  geom_point(mapping = aes(x=displ, y = hwy, color = class, shape = class,stroke=1))

ggplot(data=mpg) +
  geom_point(mapping = aes(x=displ, y = hwy, color = class, shape = class,stroke=5))

ggplot(data=mpg) +
  geom_point(mapping = aes(x=displ, y = hwy, color = class, stroke = .1),shape=22)
```

It seems as though all point shapes will take stroke as aesthetic

##### 6.  map an aesthetic to a value, not a variable
```{r}
ggplot(data=mpg) +
  geom_point(mapping = aes(x=displ, y = hwy, color = displ < 5))
```

Kind of cool.  splits into two colors, less than or greater than 5

###  3.5: Facets
facet_wrap is for one variable: facet_wrap(~ class, nrow = 3)
facet_grid is for 2 varaibles: facet_grid(drv ~ cyl)

#### 3.5.1 Exercises

##### 1. facet on a continuous variable
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = cty)) +
  facet_wrap(~ hwy)
```

This is possible, but messy.  Facets each value independently

##### 2. Why empty cells in facet_grid(drv ~ cyl)?
Missing categories.  There are no 4 cylinder, rear wheel drive cars, for example.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = cyl))
```

Overplotting, but again some categories are absent.

##### 3. What does . do in the following code?
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```

I faceted only by drive (made 3 rows)
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

This faceted only by cyl, making 4 columns.

##### 4. Look at first faceted plot in below graph
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap( ~ class, nrow = 2)

```

Advantages to faceting over color by class: can clearly see separate classes.  Might not be that useful for small datasets.
Conversely, if too many classes faceting also not useful.

##### 5. facet_wrap vs facet_grid
```{r}
?facet_wrap
# You can facet by multiple variables
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~ cyl + drv)  # cool; I didn't know that worked

# free scales can also be handy
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~class, scales = "free")

```

Handy because it wraps data into roughly symmetrical array of panels (while facet_grid(. ~ class) would instead give me a bunch of columns)

With facet_wrap, can specify number of columns (ncol) and rows (nrow) in the output array.  Note that I can also specify faceting by 2 values in facet_wrap

#####6. With facet_grid, you usually should put variable with more unique levels in columns.  Why?

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ class)
```

Given the shape of my monitor, this works pretty well
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(class ~ drv)
```

But this is an inefficient use of space - the x axes are much wider than y axes are tall.

Also, we usually display the dependent variable in the y axis (this is where we want to see differences.)  Best to keep full dynamic range there.




