---
title: "Rclub_June28_Ch13"
author: "Stacey Harmer"
date: "6/26/2017"
output: 
  html_document: 
    keep_md: yes
---

Finish chapter 13 (Relational data) 13.3 - 13.7

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 13.3  Keys

Primary key - uniquely identifies observation in its own table

Foreign key - 

```{r}
library(tidyverse)
library(nycflights13)

planes # tailnum is primary key; everything else relates back to it
length(planes$tailnum) # 3322
length(unique(planes$tailnum)) # 3322
flights # tailnum is foreign; 
length(flights$tailnum) # 336776
length(unique(flights$tailnum)) # 4044
```

I should get used to using count instead

```{r}
planes %>% 
  count(tailnum) %>% 
  filter(n > 1)  # none

flights %>% 
  count(tailnum) %>% 
  filter(n > 1)  # 3,873 × 2

weather %>% 
  count(year, month, day, hour, origin) %>% 
  filter(n > 1) # also none


```

YOu can add a primary key using 'mutate' and row_number  (this is a surrogate key)

### 13.3.1 Exercises

#### 13.3.1.1  Add a surrogate key to flights.

```{r}
flights.new.key <- flights %>%
  mutate(sur.key = row_number())

str(flights.new.key)
summary(flights.new.key)
```

#### 13.3.1.2  Identify the keys in the following datasets

```{r}
?Lahman::Batting
head(Lahman::Batting)
```

I'd guess playerID is primary key
```{r}

Lahman::Batting %>% 
  count(playerID, yearID, stint) %>% 
  filter(n > 1)  # that is unique

Lahman::Batting %>% 
  count(playerID, yearID, teamID) %>% 
  filter(n > 1)  # not unique

```

```{r}
library(babynames)
head(babynames)

# I'm guessing year, sex, name must all be combined to be unique
babynames %>%
  count(year, name) %>%
  filter(nn > 1)  #163,000 rows!

babynames %>%
 count(year, name, sex) %>%
 filter(nn > 1)  ## that did it

```


```{r}
library(nasaweather)
head(atmos)

atmos %>%
  count(lat, long, year, month) %>%
  filter(n > 1) # that is now unique

```

```{r}
library(fueleconomy)
head(vehicles)

vehicles %>%
  count(id) %>%
  filter(n > 1)  # finally!  A key variable

```

```{r}
head(diamonds)

diamonds %>%
  count(carat, cut, color) %>%
  filter( n > 1) # not unique

diamonds %>%
  count(carat, cut, color, clarity) %>%
  filter( n > 1)

diamonds %>%
  count(carat, cut, color, clarity, depth, table, price) %>%
  filter( n > 1) # still have 735 rows.  

diamonds %>%
  count(carat, cut, color, clarity, depth, table, price, x , y , z) %>%
  filter( n > 1)  # still have 143 rows.
## WHY???????

diamonds %>%
  filter(carat == 0.30, 
         cut == "Good", 
         color == "J", 
         clarity == "VS1", 
         depth == 63.4, 
         table == 57) 

# These entries look hte same to me.  Nots sure whey they are listed twice
diamonds # A tibble: 53,940 × 10
unique(diamonds) # 53,794 × 10

```

#### 13.3.1.4 Draw a diagram illustrating the connections between the Batting, Master, and Salaries tables in the Lahman package. Draw another diagram that shows the relationship between Master, Managers, AwardsManagers.

```{r}
library(Lahman)
head(Batting)  # playerID, yearID, stint, are the primary keys; then a bunch of stats for the player
head(Master)  # info on each player.  Primary key is playerID.  easy to relate to Batting
head(Managers)  # can get teamID and year here; relate back to Batting to see who the manager was.  Possible complication: can have multilple managers per year; if so, might be hard to figure out how 'stint' in Batting relates to managers
head(Salaries) # can related playerID and year and team to performance in the Batting tibble.

```

How would you characterise the relationship between the Batting, Pitching, and Fielding tables?

```{r}
?Pitching #  44139 observations 
?Batting  # 101332 observations - more people bat than pitch
?Fielding # 170526 observations - more people field than bat (?)
```

These are parallel tibbles I think.  

## 13.4  Mutating joins

TO combine variables from 2 tables.

```{r}
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)
flights2

```

Now add full airline name to the above
```{r}
head(airlines)

flights2 %>%
  select(-origin, -dest) %>% 
  left_join(airlines, by = "carrier")

```

### 13.4.1  Understanding joins

### 13.4.2 Inner join.  

Note that the output will only be the rows that match each other - danger!!

### 13.4.3 Outer join.  

Specify what is to be retained.  left_join keeps all in hte first table (x); right_join keeps all in teh second table (y); full_join keeps all observations in both tables.

Left join is default join!!!

### 13.4.4 Duplicate keys.

What happens when keys are not unique??  

1.the joining key is primary in one table but not in the other.  The information in the non-primary key is duplicated.

```{r}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2"
)
left_join(x, y, by = "key")

```

2.  Both tables have duplicate keys.  Problem!

### 13.4.5 Defininig the key columns

```{r}
head(flights2)
head(weather)

flights %>%
  left_join(weather)
```

Natural join just uses all the column names that match between the tables

### 13.4.6 Exercises

### 13.4.6.1  Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays.  

I assume this means average arrival delay (?)

```{r}
head(flights) # dest has the faa code for destination
head(airports)  # faa is the airport code

avg.delays <- flights %>%
  group_by(dest) %>%
  summarize(avg.arr.del = mean(arr_delay, na.rm = T)) %>%
  right_join(airports, c("dest" = "faa")) %>%
  filter(avg.arr.del != "NA")

avg.delays
# gives a tibble with 101 rows.
```

Now put this data onto the US map

```{r}

airports %>%
  semi_join(flights, c("faa" = "dest"))
# this didn't do what I expected

avg.delays %>%
  ggplot(aes(lon, lat, size = avg.arr.del)) +
    borders("state") +
    geom_point() +
    coord_quickmap()

avg.delays %>%
  ggplot(aes(lon, lat, color = avg.arr.del)) +
    borders("state") +
    geom_point() +
    coord_quickmap() # better

avg.delays %>%
  ggplot(aes(lon, lat, color = avg.arr.del)) +
    borders("state") +
    geom_point() +
    coord_quickmap() +
    scale_colour_gradient(low = "blue", high = "red")

```

### 13.4.6.2 Add the location of the origin and destination (i.e. the lat and lon) to flights.

This would be an left join
```{r}
head(flights)
head(airports)

flights %>%
  left_join(airports, by = c("origin" = "faa")) %>%
  left_join(airports, by = c("dest" = "faa"))

#works, but very messy.
```

### 13.4.6.3 Is there a relationship between the age of a plane and its delays?

To address this, I will join flights with planes

```{r}
head(flights)
head(planes)

flights.age <- 
  flights %>%
  left_join(planes, by = "tailnum")

cor.test(flights.age$arr_delay, flights.age$year.y)
# cor is 0.01767153; nope, not much correlation

ggplot(flights.age, aes(x = year.y, y = arr_delay)) +
  geom_hex()

```

The one striking correlation is that htere are more delays as the years go on.

BElow is code from answer key
```{r}
plane_ages <- 
  planes %>%
  mutate(age = 2017 - year) %>%
  select(tailnum, age)

flights %>%
  inner_join(plane_ages, by = "tailnum") %>%
  group_by(age) %>%
  filter(!is.na(dep_delay)) %>%
  summarise(delay = mean(dep_delay)) %>%
  ggplot(aes(x = age, y = delay)) +
  geom_point() +
  geom_line()

```

### 13.4.6.4  What weather conditions make it more likely to see a delay?

```{r}
head(weather)
head(flights)

# I am just going to consider dep_delay for this one

flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), temp = mean(temp, na.rm = T)) %>%
  ggplot(aes(x = temp, y = delay)) +
  geom_point()

# there seems to be a correlation!  Is this just time of day?

flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), temp = mean(temp, na.rm = T)) %>%
  ggplot(aes(x = temp, y = hour)) +
  geom_point()

flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), temp = mean(temp, na.rm = T)) %>%
  ggplot(aes(x = hour, y = delay)) +
  geom_point()  # for sure a correlation here (dep hour and delay)


flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), visib = mean(visib, na.rm = T)) %>%
  ggplot(aes(x = visib, y = delay)) +
  geom_point() # looks like visib is related to temp


flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), visib = mean(visib, na.rm = T),
            temp = mean(temp, na.rm = T)) %>%
  ggplot(aes(x = visib, y = temp)) +
  geom_point()  #these seem correlated

# wind

flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), visib = mean(visib, na.rm = T),
            temp = mean(temp, na.rm = T), wind = mean(wind_speed, na.rm = T)) %>%
  ggplot(aes(x = wind, y = delay)) +
  geom_point()  #these seem correlated also - maybe best yet

# gusts
flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), visib = mean(visib, na.rm = T),
            temp = mean(temp, na.rm = T), wind.gust = mean(wind_gust, na.rm = T)) %>%
  ggplot(aes(x = wind.gust, y = delay)) +
  geom_point()  # similar to above

# time of day component with wind?
flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), visib = mean(visib, na.rm = T),
            temp = mean(temp, na.rm = T), wind.gust = mean(wind_gust, na.rm = T)) %>%
  ggplot(aes(x = hour, y = wind.gust)) +
  geom_point()  # yes, these are correlated too.


# rain

flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(c(year:dep_delay, temp:visib, hour)) %>%
  group_by(hour) %>%
  summarize(delay = mean(dep_delay, na.rm = T), visib = mean(visib, na.rm = T),
            temp = mean(temp, na.rm = T), precip = mean(precip, na.rm = T)) %>%
  ggplot(aes(x = precip, y = delay)) +
  geom_point()  # not a great correlation here
```

Looks to me like time of day matters a lot, as does wind speed, wind gusts, and maybe visibility
However, weather is correlated with time of day, so hard to say which matter most.

### 13.4.6.5  What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.

First, need to figure out delays on this day (by origin and by hour)
```{r}

# Want to first filter flights so I have only that day

flights.June13.2013 <- flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
# group by origin and by hour
  group_by(origin, hour) %>% # and now calculate mean departure delays
  summarize(delay = mean(dep_delay, na.rm = T)) %>%
  left_join(airports, by = c("origin" = "faa"))
  

flights.June13.2013 %>%
  ggplot(aes(lon, lat, color = delay)) +
    borders(database = "state", region = c("new york")) +
    geom_point() +
    scale_colour_gradient(low = "blue", high = "red") +
    coord_quickmap() # better, but I'd like to just see the city.

# not a great display
```

Google says: An unusually strong non-tropical low pressure for June affected the Northeast on June 13-14, 2013, following tropical storm Andrea and another low pressure earlier in the same week that brought Central Park to near record June rainfall. While impacts from this storm were less significant than originally expected, it managed to produce another 1-3 inches of rain in the Northeast and a widespread severe weather outbreak in the Mid Atlantic region.

The answer key suggested the following (looking at arrivals at dest!):

```{r}
library(viridis)
flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c("dest" = "faa")) %>%
  ggplot(aes(y = lat, x = lon, size = delay, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap() + 
  scale_color_viridis()

# OK, that is better than mine
```

Might have been interesting to plot departure delay by time across the day

```{r}
flights.June13.2013  %>%
  ggplot(aes(x = hour, y = delay)) +
  geom_point() + 
  facet_grid( origin ~ .)
```

A bad day to travel

## 13.5 Filtering joins

These affect the obsevations, not the variables
```{r}
top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest

```
semi-join keeps ony the rows in x that have a match in y

```{r}
flights %>% 
  semi_join(top_dest)

flights %>% 
  semi_join(top_dest) %>%
  distinct(dest) # indeed, only 10 airports
```

I don't really understand this well

```{r}
x <- tribble(
  ~key, ~val_x, ~gt,
     1, "x1", "Col",
     2, "x2", "Ler",
     3, "x3", "Col"
)
x
y <- tribble(
  ~key, ~val_y, ~gt,
     1, "y1", "Col",
     2, "y2", "Col",
     4, "y3", "WS"
)
y
semi_join(x, y)

```


### 13.5.1 Exercises

#### 13.5.1.1  What does it mean for a flight to have a missing tailnum? What do the tail numbers that don’t have a matching record in planes have in common? (Hint: one variable explains ~90% of the problems.)

```{r}
summary(planes) # all planes in the tibble have a tailnumber, so he must be asking about the join between the flights and the planes tibbles

summary(flights) # there is a tailnum for every entry, so the issue is that some of the tailnum in flights are not present in the planes tibble

flights %>%
  anti_join(planes, by = "tailnum") %>%
  select(time_hour, origin, dest, carrier) %>%
  distinct(carrier)
# 10 carriers

flights %>%
  anti_join(planes, by = "tailnum") %>%  # 52,606 flights without recognized tailnum
  select(time_hour, origin, dest, carrier) %>%
  count(carrier, sort = T)

flights %>%
  count(carrier, sort = T)
# I can see that few MQ flights have a recognized tailnum in the planes tibble
# AA isn't much beter
(25397 + 22558) /52606  # = 0.911588

```

#### 13.5.1.2  Filter flights to only show flights with planes that have flown at least 100 flights.

I guess this means only show flights where the total # of flights with specific tailnum is over 100

```{r}

freq.flyers <- flights %>%
  count(tailnum, sort = TRUE) %>%
  filter(n > 100)

summary(freq.flyers) # seems fine

flights %>%
  semi_join(freq.flyers)

```

#### 13.5.1.3  Combine fueleconomy::vehicles and fueleconomy::common to find only the records for the most common models.

```{r}
library(fueleconomy)
head(vehicles) #33,442 rows
head(common) #347 common models

vehicles %>%
  semi_join(common, by = c( "make","model") ) #14,531 rows

```

#### 13.5.1.4  Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the weather data. Can you see any patterns?

The weather data is only for LGA, JFK, and EWR, so I guess the rlevant delays are dep_delay

How can I select for 48 hour windows?  I could calculate mean delay for each year, month,day, hour and then somehow do a sliding window over 48 rows.

"If you need rolling aggregates (i.e. a sum computed over a rolling window), try the RcppRoll package.""
```{r}
library(RcppRoll) # roll_mean looks useful

# First, calculate mean delay for each hour

mean.delays <- flights %>%
  group_by(year, month, day, hour) %>%
summarize(mean_delay = mean(dep_delay, na.rm = T))

mean.delays #6,936 rows
summary(mean.delays)

rolling_means <- mean.delays %>%
  ungroup() %>%
  mutate(rolling_mean_48 = roll_mean(mean_delay, n = 48, na.rm = T, fill = NA, align = "right"))

summary(rolling_means)

rolling_means %>%
  filter(rolling_mean_48 == max(rolling_mean_48, na.rm = T))
```

OK!  2013, June 28.  Google doesn't reveal much on that date.

Now crosss-reference with the weather data

```{r}
weather %>%
  semi_join(
    rolling_means %>%
  filter(rolling_mean_48 == max(rolling_mean_48, na.rm = T))
  )
# no obvious problems on that day. Expand to a few days around it

View(weather %>%
  filter(year == 2013, month == 6, day == 25:28))
```
The weather still seems fine.

#### 13.5.1.5  What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?

```{r}
head(flights)
head(airports)

anti_join(flights, airports, by = c("dest" = "faa"))
# these would be flights going to airports not in the airports tibble

anti_join(airports, flights, by = c("faa" = "dest"))
# these are airports listed in the airports tibble but that aren't visited by flights listed in flight tibble.

```

#### 13.5.1.6  You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.

```{r}
head(airlines) # list of carrier names
head(planes) # tailnum is listed
head(flights) # tailnum and carrier is listed.

# So I think the question is whether the same tailnum is always linked with the same carrier in the flights tibble

# I think I want to merge flights with itself using tailnum as the key, and then compare carrier.x with carrier.y.  Inner join?

not.matched <- inner_join(flights, flights, by = "tailnum") %>%  # generates a very long table, with 37 columns and 63,032,928 rows
  filter(carrier.x != carrier.y)

# the full code returns tibble that is  4,318,330 × 37 - much too big

# let's hone in on what is of interest
not.matched %>%
  filter(tailnum != "NA") %>%
  select(tailnum, carrier.x, carrier.y) %>%
  distinct()  # this gives me 34 rows

# how many planes?
not.matched %>%
  filter(tailnum != "NA") %>%
  select(tailnum) %>%
  distinct()  # 17 tailnum


View(not.matched %>%
  filter(tailnum != "NA") %>%
  select(tailnum, carrier.x, carrier.y) %>%
  distinct()  )

```

I see same tailnum shared by FL (AirTran) and DL (Delta), by 9E (Endeavor) and EV (ExpressJet)
Googling tells me AirTran was acquired by Sowthwest in 2014, and in 2013 Endeavor became a subsidiary of Delta.

The answer key is more elegant:
```{r}
flights %>%
  group_by(tailnum, carrier) %>%
  count() %>%
  filter(n() > 1) %>%
  select(tailnum) %>%
  distinct()  # 18 tailnumbers

```

